---
title: 'TMA4315: Project 1'
author: "Jim Totland, Martin Gudahl Tufte"
date: "9/8/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\textcolor{red}{Litt usikker på hva slags notasjon vi skal bruke, f. eks. boldface for vektorer eller ikke? Bare si ifra hvis du vil ha noe spesifikit:)}

# Problem 1
## a)
Since the response variables $y_i \sim \mathrm{Bernoulli}(p_i)$, where $p_i = \Pr(y_i=1 \mid x_i) = \Phi(x_i^T \beta)$, the conditional mean is given by $\mathrm{E}y_i = p_i$, which is connected to the covariates via the following relationship:

$$
x_i^T\beta =: \eta_i = \Phi^{-1}(p_i),
$$
which implies that $p_i = \Phi(\eta_i)$. This results in the likelihood function

$$
\begin{split}
L(\beta) &= \prod_{i= 1}^n p_i^{y_i}(1-p_i)^{1-y_i} \\
&=\prod_{i = 1}^n\Phi(\eta_i)^{y_i}(1 - \Phi(\eta_i))^{1-y_i}.
\end{split}
$$
Thus, the log-likelihood becomes

$$
l(\beta) := \mathrm{ln}(L(\beta)) = \sum_{i = 1}^n y_i \mathrm{ln}(\Phi(\eta_i)) + (1- y_i)\mathrm{ln}(1-\Phi(\eta_i)) = \sum_{i = 1}^nl_i(\beta).
$$
To find the score function, we calculate

$$
\begin{split}
  \frac{\partial l_i(\beta)}{\partial \beta} &= \frac{y_i}{\Phi(\eta_i)} \frac{\partial
  \Phi(\eta_i)}{\partial   \beta} - \frac{1-y_i}{1-\Phi(\eta_i)} \frac{\partial \Phi(\eta_i)}{\partial \beta} \\
  &= \frac{y_i}{\Phi(\eta_i)} \phi(\eta_i)x_i - \frac{1-y_i}{1-\Phi(\eta_i)} \phi(\eta_i)x_i \\
  &= \frac{y_i(1-\Phi(\eta_i)) - (1-y_i)\Phi(\eta_i)}{\Phi(\eta_i)(1-\Phi(\eta_i))}\phi(\eta_i)x_i \\
  &= \frac{y_i - \Phi(\eta_i)}{\Phi(\eta_i)(1-\Phi(\eta_i))}\phi(\eta_i)x_i.
\end{split}
$$
Consequently, the score function is given by 

$$
s(\beta ) = \sum_{i = 1}^n \frac{y_i - \Phi(\eta_i)}{\Phi(\eta_i)(1-\Phi(\eta_i))}\phi(\eta_i)x_i.
$$

Next, we find the expected Fisher information, $F(\beta)$. We find it by using the result

$$
\begin{split}
  F(\beta) &= \mathrm{Var}(s(\beta)) = \mathrm{Var} \left( \sum_{i = 1}^n \frac{y_i - \Phi(\eta_i)}{\Phi(\eta_i)(1-\Phi(\eta_i))}\phi(\eta_i)x_i \right) \\
  &= \sum_{i = 1}^n \underbrace{\left[ \frac{\phi(\eta_i)}{\Phi(\eta_i)(1 - \Phi(\eta_i))}\right]^2}_{ =: \xi_i} \mathrm{Var}(y_ix_i) = \sum_{i = 1}^n \xi_i x_i \mathrm{Var}(y_i) x_i^T \\
  &= \sum_{i = 1}^n \xi_i p_i(1-p_i) x_i x_i^T \\
  &= \sum_{i = 1}^n \frac{\phi(\eta_i)^2}{\Phi(\eta_i)(1 - \Phi(\eta_i))} x_i x_i^T,
\end{split}
$$
Where in the third equality we have used that the $y_i$'s are independent. The expected Fisher information can also be verified to have this expression by the relationship
$$
  F(\beta) = \sum_{i=1}^n \frac{h'(\eta_i)^2}{\mathrm{Var}(y_i)} x_i x_i^T,
$$
where $h'(\eta_i) = \Phi'(\eta_i) = \phi(\eta_i)$ and $\mathrm{Var}(y_i) = p_i(1-p_i) = \Phi(\eta_i)(1-\Phi(\eta_i))$.

 
## b) The expected Fisher information is given by

$$
  F(\beta) = \sum_{i = 1}^n \frac{\phi(\eta_i)^2}{\Phi(\eta_i)(1 - \Phi(\eta_i))} x_i x_i^T = x^T W x,
$$
where $W = \mathrm{diag}\left(\frac{\phi(\eta_i)^2}{\Phi(\eta_i)(1 - \Phi(\eta_i))}\right)$.

The Fisher scoring algorithm states that the next iterate is given by

$$
  \beta^{(t+1)} = \beta^{(t)} + F(\beta(t))^{-1} s(\beta(t)).
$$
Inserting the expected Fisher information and the score function we get

$$
  \beta^{(t+1)} = (x^T W^{(t)} x)^{-1} x^T W^{(t)} \tilde y^{(t)},
$$
where the working response vector $\tilde y^{(t)}$ has element $i$ given by

$$
  \tilde y_i^{(t)} = x_i^T \beta^{(t)} + \frac{y_i - h(x_i^T \beta(t))}{h'(x_i^T \beta(t))} = \eta_i^{(t)} + \frac{y_i - \Phi(\eta_i^{(t)})}{\phi(\eta_i^{(t)})}.
$$
Implementing myglm in R:

```{r}
Phi <- function(x) return (pnorm(x))
phi <- function(x) return (dnorm(x))


myglm <- function(formula, data, start = NULL){
  
  # response variable
  resp <- all.vars(formula)[1]
  y <- as.matrix( data[resp] )
  
  
  # model matrix
  X <- model.matrix(formula, data)
  n <- dim(X)[1]
  p <- dim(X)[2]
  
  
  # starting beta
  if (is.null(start)){
    beta = rep(0, p)
  }
  else {
    beta = start
  }
  
  
  # Fisher scoring algorithm
  max_iter <- 50
  tol <- 1e-10
  iter <- 0
  rel.err <- Inf
  
  while (rel.err > tol & iter < max_iter){
    # calculate eta, y tilde, W
    eta <- X %*% beta
    y.tilde <- eta + (y - Phi(eta)) / (phi(eta))
    W <- diag( as.vector(phi(eta)^2 / (Phi(eta)*(1-Phi(eta)))), n, n )
    
    
    # update beta
    A <- t(X) %*% W %*% X
    b <- t(X) %*% W %*% y.tilde
    beta.new <- solve(A, b)
    
    iter <- iter + 1
    rel.err <- max(abs(beta.new - beta) / abs(beta.new))
    beta <- beta.new
  }
  
  
  
  
  # remains to find the coefficients matrix, deviance and estimated variance matrix
  
  coeff <- 1
  deviance <- 1
  vcov <- 1
  
  return (beta)
}

beta <- myglm(menarche ~ age, juul.girl)

beta

#X <- model.matrix(menarche ~ age, juul.girl)

```

## c)

```{r}

# probability
x = runif(1000, 0, 1)
# draw n bernoulli with prob x
y <- rbinom(1000, 1, x)

df <- data.frame(y, x)



### fit using glm
model <- glm(y ~ poly(x,2), family = binomial(link = "probit"), data = df)

# beta
model$coefficients

# vcov
vcov(model)

# deviance
# ...

### fit using myglm
beta <- myglm(y ~ poly(x,2), data = df)

# beta
t(beta)

# vcov
# ...

# deviance
# ...

```




 
 
# Problem 2
## a)
 
```{r}
#install.packages("ISwR")
library(ISwR) # Install the package if needed
data(juul)
juul$menarche <- juul$menarche - 1
juul.girl <- subset(juul, age>8 & age<20 & complete.cases(menarche))
```
 
 
```{r}
model <- glm(menarche ~ age, family=familytype(link="probit"), data= juul.girl)
```

